---
title: "ES3802 Air Quality Analysis"
author: "Fun Jolie Wants To Die"
date: "2025-03-30"
output:
  pdf_document:
    toc: true
    toc_depth: '3'
  html_document:
    code_folding: hide
    toc: true
    toc_depth: 3
    theme: united
---

# Data cleaning & Arranging

Load required packages
```{r, warning=FALSE, message=FALSE}
#Install packages as needed, honestly you're a noob if you dont own these
library(ggplot2)
library(gridExtra)
library(tidyverse) #for dyplr
library(corrplot) #correlation plot
library(readxl)
```

Load data
```{r}
#NOTE: setwd NOT specified, just make sure the file is located the same as the code
data = read_xlsx("ES3802_Data.xlsx")

#remove 2nd row, its units
data = data[-1,]

#make sure R recognize it as a numeric
data$PN <- as.numeric(as.character(data$PN))
```

Arrange data base on point number in morning and afternoon
```{r}
# Arrange by Session and then Point Number

data = data %>%
  arrange(Session, PN)

#Rename points  to new point number
data = data %>%
  mutate(PN = case_when(
    PN == 8  ~ 6,
    PN == 13 ~ 10,
    PN == 17 ~ 14,
    PN == 20 ~ 17,
    PN == 22 ~ 19,
    PN == 24 ~ 21,
    PN == 26 ~ 23,
    PN == 29 ~ 26,
    PN == 30 ~ 27,
    PN == 27 ~ 24,
    PN == 28 ~ 25,
    PN == 25 ~ 22,
    PN == 23 ~ 20,
    PN == 21 ~ 18,
    PN == 19 ~ 16,
    PN == 16 ~ 13,
    PN == 12 ~ 9,
    PN == 7  ~ 5,
    PN == 3  ~ 2,
    PN == 6  ~ 4,
    PN == 11 ~ 8,
    PN == 15 ~ 12,
    PN == 18 ~ 15,
    PN == 14 ~ 11,
    PN == 10 ~ 7,
    PN == 5  ~ 3,
    PN == 2  ~ 1,
  ))
```

# Calculate Average, Min & Max
```{r}
# Initialize an empty data frame to store results
result = data.frame(Session = numeric(0), PN = numeric(0), 
                    Min_PM1 = numeric(0), Max_PM1 = numeric(0), Avg_PM1 = numeric(0),
                    Min_PM2.5 = numeric(0), Max_PM2.5 = numeric(0), Avg_PM2.5 = numeric(0),
                    Min_RESP = numeric(0), Max_RESP = numeric(0), Avg_RESP = numeric(0),
                    Min_PM10 = numeric(0), Max_PM10 = numeric(0), Avg_PM10 = numeric(0),
                    Min_TOTAL = numeric(0), Max_TOTAL = numeric(0), Avg_TOTAL = numeric(0))

# Ensure the required columns are numeric (coerce them if necessary)
data$PM1 = as.numeric(as.character(data$PM1))
data$PM2.5 = as.numeric(as.character(data$PM2.5))
data$RESP = as.numeric(as.character(data$RESP))
data$PM10 = as.numeric(as.character(data$PM10))
data$TOTAL = as.numeric(as.character(data$TOTAL))

# Get unique sessions and point numbers
sessions = unique(data$Session)
point_numbers = unique(data$PN)

# Loop through each session and point number
for (session in sessions) {
  for (pn in point_numbers) {
    # Subset data for current session and point number
    subset_data = data[data$Session == session & data$PN == pn, ]
    
    # Calculate min, max, and average for each column
    min_pm1 = min(subset_data$PM1, na.rm = TRUE)
    max_pm1 = max(subset_data$PM1, na.rm = TRUE)
    avg_pm1 = mean(subset_data$PM1, na.rm = TRUE)
    
    min_pm25 = min(subset_data$PM2.5, na.rm = TRUE)
    max_pm25 = max(subset_data$PM2.5, na.rm = TRUE)
    avg_pm25 = mean(subset_data$PM2.5, na.rm = TRUE)
    
    min_resp = min(subset_data$RESP, na.rm = TRUE)
    max_resp = max(subset_data$RESP, na.rm = TRUE)
    avg_resp = mean(subset_data$RESP, na.rm = TRUE)
    
    min_pm10 = min(subset_data$PM10, na.rm = TRUE)
    max_pm10 = max(subset_data$PM10, na.rm = TRUE)
    avg_pm10 = mean(subset_data$PM10, na.rm = TRUE)
    
    min_total = min(subset_data$TOTAL, na.rm = TRUE)
    max_total = max(subset_data$TOTAL, na.rm = TRUE)
    avg_total = mean(subset_data$TOTAL, na.rm = TRUE)
    
    # Append the results to the result data frame
    result = rbind(result, data.frame(
      Session = session, PN = pn, 
      Min_PM1 = min_pm1, Max_PM1 = max_pm1, Avg_PM1 = avg_pm1,
      Min_PM2.5 = min_pm25, Max_PM2.5 = max_pm25, Avg_PM2.5 = avg_pm25,
      Min_RESP = min_resp, Max_RESP = max_resp, Avg_RESP = avg_resp,
      Min_PM10 = min_pm10, Max_PM10 = max_pm10, Avg_PM10 = avg_pm10,
      Min_TOTAL = min_total, Max_TOTAL = max_total, Avg_TOTAL = avg_total
    ))
  }
}
```

# Data exploration 

Histogram of average values
```{r}
Hist_Avg_PM1 = ggplot(result, aes(x = Avg_PM1)) +
  geom_histogram(binwidth = 0.001, fill = "lightblue", color = "black", alpha = 0.7) +
  labs(title = "Histogram of Avg_PM1", x = "Avg_PM1", y = "Count") +
  theme_minimal()

Hist_Avg_PM2.5 = ggplot(result, aes(x = Avg_PM2.5)) +
  geom_histogram(binwidth = 0.001, fill = "lightblue", color = "black", alpha = 0.7) +
  labs(title = "Histogram of Avg_PM2.5", x = "Avg_PM2.5", y = "Count") +
  theme_minimal()

Hist_Avg_PM10 = ggplot(result, aes(x = Avg_PM10)) +
  geom_histogram(binwidth = 0.001, fill = "lightblue", color = "black", alpha = 0.7) +
  labs(title = "Histogram of Avg_PM1", x = "Avg_PM1", y = "Count") +
  theme_minimal()

Hist_Avg_RESP = ggplot(result, aes(x = Avg_RESP)) +
  geom_histogram(binwidth = 0.001, fill = "lightblue", color = "black", alpha = 0.7) +
  labs(title = "Histogram of Avg_PM2.5", x = "Avg_PM2.5", y = "Count") +
  theme_minimal()

Hist_Avg_TOTAL = ggplot(result, aes(x = Avg_TOTAL)) +
  geom_histogram(binwidth = 0.001, fill = "lightblue", color = "black", alpha = 0.7) +
  labs(title = "Histogram of Avg_PM1", x = "Avg_PM1", y = "Count") +
  theme_minimal()

print(Hist_Avg_PM1)
print(Hist_Avg_PM2.5)
print(Hist_Avg_PM10)
print(Hist_Avg_RESP)
print(Hist_Avg_TOTAL)

```
Data is NOT normally distributed, and there is some outliers ~0.04 (yay david)

# Data Analysis
Is there a difference between in the average value for each point number between the 2 session?

Use a for loop to iterate over each parameter and perform the paired t-test for point
```{r}
S1_PM1 = result[result$Session == 1, "Avg_PM1"]
S2_PM1 = result[result$Session == 2, "Avg_PM1"]

S1_PM2.5 = result[result$Session == 1, "Avg_PM2.5"]
S2_PM2.5 = result[result$Session == 2, "Avg_PM2.5"]

S1_PM10 = result[result$Session == 1, "Avg_PM10"]
S2_PM10 = result[result$Session == 2, "Avg_PM10"]

S1_RESP = result[result$Session == 1, "Avg_RESP"]
S2_RESP = result[result$Session == 2, "Avg_RESP"]

S1_TOTAL = result[result$Session == 1, "Avg_TOTAL"]
S2_TOTAL = result[result$Session == 2, "Avg_TOTAL"]

ttest_PM1 = t.test(S1_PM1, S2_PM1)
ttest_PM2.5 = t.test(S1_PM2.5, S2_PM2.5)
ttest_PM10 = t.test(S1_PM10, S2_PM10)
ttest_RESP = t.test(S1_RESP, S2_RESP)
ttest_TOTAL = t.test(S1_TOTAL, S2_TOTAL)

print(ttest_PM1)
print(ttest_PM2.5)
print(ttest_PM10)
print(ttest_RESP)
print(ttest_TOTAL)
```
Welch two sample t-test shows that there is no difference in the values between the parameters taken in the morning and in the afternoon



Plot for visualisation
```{r}
# Combine the data into a single data frame
PM1 = data.frame(
  Point = rep(1:length(S1_PM1), 2),  # Repeat the point numbers for both sessions
  PM1 = c(S1_PM1, S2_PM1),           # Combine S1 and S2 PM1 values
  Session = rep(c("Session 1", "Session 2"), each = length(S1_PM1))  # Label the sessions
)

# Create the ggplot
PM1_plot = ggplot(PM1, aes(x = Point, y = PM1, color = Session)) +
  geom_point() +                      # Add points
  geom_line() +                       # Add lines connecting the points
  labs(x = "Point Number", y = "PM1 Values", title = "Comparison of PM1 for Session 1 and Session 2") +
  scale_color_manual(values = c("blue", "red")) +  # Custom colors for each session
  theme_minimal()         

# For PM2.5
PM2.5 <- data.frame(
  Point = rep(1:length(S1_PM2.5), 2),  # Repeat the point numbers for both sessions
  PM2.5 = c(S1_PM2.5, S2_PM2.5),       # Combine S1 and S2 PM2.5 values
  Session = rep(c("Session 1", "Session 2"), each = length(S1_PM2.5))  # Label the sessions
)

# Create the ggplot for PM2.5
PM2.5_plot <- ggplot(PM2.5, aes(x = Point, y = PM2.5, color = Session)) +
  geom_point() +                         # Add points
  geom_line() +                          # Add lines connecting the points
  labs(x = "Point Number", y = "PM2.5 Values", title = "Comparison of PM2.5 for Session 1 and Session 2") +
  scale_color_manual(values = c("blue", "red")) +  # Custom colors for each session
  theme_minimal()                         # Clean theme

# For PM10
PM10 <- data.frame(
  Point = rep(1:length(S1_PM10), 2),  # Repeat the point numbers for both sessions
  PM10 = c(S1_PM10, S2_PM10),         # Combine S1 and S2 PM10 values
  Session = rep(c("Session 1", "Session 2"), each = length(S1_PM10))  # Label the sessions
)

# Create the ggplot for PM10
PM10_plot <- ggplot(PM10, aes(x = Point, y = PM10, color = Session)) +
  geom_point() +                         # Add points
  geom_line() +                          # Add lines connecting the points
  labs(x = "Point Number", y = "PM10 Values", title = "Comparison of PM10 for Session 1 and Session 2") +
  scale_color_manual(values = c("blue", "red")) +  # Custom colors for each session
  theme_minimal()                         # Clean theme

# For RESP
RESP <- data.frame(
  Point = rep(1:length(S1_RESP), 2),  # Repeat the point numbers for both sessions
  RESP = c(S1_RESP, S2_RESP),         # Combine S1 and S2 RESP values
  Session = rep(c("Session 1", "Session 2"), each = length(S1_RESP))  # Label the sessions
)

# Create the ggplot for RESP
RESP_plot <- ggplot(RESP, aes(x = Point, y = RESP, color = Session)) +
  geom_point() +                         # Add points
  geom_line() +                          # Add lines connecting the points
  labs(x = "Point Number", y = "RESP Values", title = "Comparison of RESP for Session 1 and Session 2") +
  scale_color_manual(values = c("blue", "red")) +  # Custom colors for each session
  theme_minimal()                         # Clean theme

# For TOTAL
TOTAL <- data.frame(
  Point = rep(1:length(S1_TOTAL), 2),  # Repeat the point numbers for both sessions
  TOTAL = c(S1_TOTAL, S2_TOTAL),       # Combine S1 and S2 TOTAL values
  Session = rep(c("Session 1", "Session 2"), each = length(S1_TOTAL))  # Label the sessions
)

# Create the ggplot for TOTAL
TOTAL_plot <- ggplot(TOTAL, aes(x = Point, y = TOTAL, color = Session)) +
  geom_point() +                         # Add points
  geom_line() +                          # Add lines connecting the points
  labs(x = "Point Number", y = "TOTAL Values", title = "Comparison of TOTAL for Session 1 and Session 2") +
  scale_color_manual(values = c("blue", "red")) +  # Custom colors for each session
  theme_minimal()                         # Clean theme
# Clean theme
```


Print and save the plots
```{r}
print(PM1_plot)
ggsave("PM1.png", plot = last_plot(), width = 8, height = 6, dpi = 300)
print(PM2.5_plot)
ggsave("PM2.5.png", plot = last_plot(), width = 8, height = 6, dpi = 300)
print(PM10_plot)
ggsave("PM10.png", plot = last_plot(), width = 8, height = 6, dpi = 300)
print(RESP_plot)
ggsave("RESP.png", plot = last_plot(), width = 8, height = 6, dpi = 300)
print(TOTAL_plot)
ggsave("TOTAL.png", plot = last_plot(), width = 8, height = 6, dpi = 300)

```
## ANOVA: significant difference in PM by distance (S1) 
See if there is a significant difference in PM in distance from source in session 1
```{r}
#row 1 is closest to source, row 4 is furthest from source

# Find the average values for specific points in Session 1
row1 = c(6, 12, 14, 17, 19)

# Use the subset to filter the rows for Session 1 and the specific points
row1data = result[result$Session == 1 & result$PN %in% row1, ]

# Calculate the mean for each parameter (Avg_PM1, Avg_PM2.5, Avg_PM10, Avg_RESP, Avg_TOTAL)
avg_values1 = colMeans(row1data[, c("Avg_PM1", "Avg_PM2.5", "Avg_PM10", "Avg_RESP", "Avg_TOTAL")])

# Repeat for other rows

# Find the average values for specific points in Session 1
row2 = c(5, 9, 13, 16)

# Use the subset to filter the rows for Session 1 and the specific points
row2data = result[result$Session == 1 & result$PN %in% row2, ]

# Calculate the mean for each parameter (Avg_PM1, Avg_PM2.5, Avg_PM10, Avg_RESP, Avg_TOTAL)
avg_values2 = colMeans(row2data[, c("Avg_PM1", "Avg_PM2.5", "Avg_PM10", "Avg_RESP", "Avg_TOTAL")])


# Find the average values for specific points in Session 1
row3 = c(2, 4, 8, 12)

# Use the subset to filter the rows for Session 1 and the specific points
row3data = result[result$Session == 1 & result$PN %in% row3, ]

# Calculate the mean for each parameter (Avg_PM1, Avg_PM2.5, Avg_PM10, Avg_RESP, Avg_TOTAL)
avg_values3 = colMeans(row3data[, c("Avg_PM1", "Avg_PM2.5", "Avg_PM10", "Avg_RESP", "Avg_TOTAL")])

# Find the average values for specific points in Session 1
row4 = c(1, 3, 7)

# Use the subset to filter the rows for Session 1 and the specific points
row4data = result[result$Session == 1 & result$PN %in% row4, ]

# Calculate the mean for each parameter (Avg_PM1, Avg_PM2.5, Avg_PM10, Avg_RESP, Avg_TOTAL)
avg_values4 = colMeans(row4data[, c("Avg_PM1", "Avg_PM2.5", "Avg_PM10", "Avg_RESP", "Avg_TOTAL")])
```

```{r}
# Combine the data for the different rows into a single data frame for analysis
# Create a new column "Row" to indicate the distance category (Row 1, Row 2, etc.)

# Combine row data into one dataframe for analysis
row_data <- rbind(
  data.frame(Distance = "Row 1", result[result$Session == 1 & result$PN %in% c(6, 12, 14, 17, 19), 
                                        c("Avg_PM1", "Avg_PM2.5", "Avg_PM10", 
                                          "Avg_RESP", "Avg_TOTAL")]),
  
  data.frame(Distance = "Row 2", result[result$Session == 1 & result$PN %in% c(5, 9, 13, 16), 
                                        c("Avg_PM1", "Avg_PM2.5", "Avg_PM10", 
                                          "Avg_RESP", "Avg_TOTAL")]),
  
  data.frame(Distance = "Row 3", result[result$Session == 1 & result$PN %in% c(2, 4, 8, 12), 
                                        c("Avg_PM1", "Avg_PM2.5", "Avg_PM10", 
                                          "Avg_RESP", "Avg_TOTAL")]),
  
  data.frame(Distance = "Row 4", result[result$Session == 1 & result$PN %in% c(1, 3, 7), 
                                        c("Avg_PM1", "Avg_PM2.5", "Avg_PM10", 
                                          "Avg_RESP", "Avg_TOTAL")])
)

# Run ANOVA for each parameter (PM1, PM2.5, PM10, RESP, TOTAL)
params <- c("Avg_PM1", "Avg_PM2.5", "Avg_PM10", "Avg_RESP", "Avg_TOTAL")
anova_results <- list()

for (param in params) {
  # Perform ANOVA for each parameter
  anova_result <- aov(get(param) ~ Distance, data = row_data)
  
  # Store the result
  anova_results[[param]] <- summary(anova_result)
}

# Print the results of the ANOVA tests
for (param in names(anova_results)) {
  cat(paste("ANOVA results for", param, ":\n"))
  print(anova_results[[param]])
  cat("\n")
}
```
Avg_PM1:
p-value = 0.00213 (Significant at the 0.01 level). There is a statistically 
significant difference in the mean PM1 values across the different distances 
from the source (Row 1, Row 2, Row 3, Row 4).

Avg_PM2.5 = 
p-value = 0.00192 (Significant at the 0.01 level). There is a statistically 
significant difference in the mean PM2.5 values across the different distances
from the source.

Avg_PM10 =
p-value = 0.00123 (Significant at the 0.01 level). There is a statistically
significant difference in the mean PM10 values across the different distances
from the source.

Avg_RESP:
p-value = 0.00133 (Significant at the 0.01 level). There is a statistically
significant difference in the mean RESP values across the different distances
from the source.

Avg_TOTAL:
p-value = 0.0719 (Not significant at the 0.05 level). There is no statistically
significant difference in the mean TOTAL values across the different distances
from the source.
The result is marginally significant at the 0.1 level (p = 0.0719).

## ANOVA: significant difference in PM by distance (S2) 
```{r}
# Find the average values for specific points in Session 2
row1 = c(6, 12, 14, 17, 19)

# Use the subset to filter the rows for Session 2 and the specific points
row1data = result[result$Session == 2 & result$PN %in% row1, ]

# Calculate the mean for each parameter (Avg_PM1, Avg_PM2.5, Avg_PM10, Avg_RESP, Avg_TOTAL)
avg_values1 = colMeans(row1data[, c("Avg_PM1", "Avg_PM2.5", "Avg_PM10", "Avg_RESP", "Avg_TOTAL")])

# Repeat for other rows

# Find the average values for specific points in Session 2
row2 = c(5, 9, 13, 16)

# Use the subset to filter the rows for Session 2 and the specific points
row2data = result[result$Session == 2 & result$PN %in% row2, ]

# Calculate the mean for each parameter
avg_values2 = colMeans(row2data[, c("Avg_PM1", "Avg_PM2.5", "Avg_PM10", "Avg_RESP", "Avg_TOTAL")])

# Find the average values for specific points in Session 2
row3 = c(2, 4, 8, 12)

# Use the subset to filter the rows for Session 2 and the specific points
row3data = result[result$Session == 2 & result$PN %in% row3, ]

# Calculate the mean for each parameter
avg_values3 = colMeans(row3data[, c("Avg_PM1", "Avg_PM2.5", "Avg_PM10", "Avg_RESP", "Avg_TOTAL")])

# Find the average values for specific points in Session 2
row4 = c(1, 3, 7)

# Use the subset to filter the rows for Session 2 and the specific points
row4data = result[result$Session == 2 & result$PN %in% row4, ]

# Calculate the mean for each parameter
avg_values4 = colMeans(row4data[, c("Avg_PM1", "Avg_PM2.5", "Avg_PM10", "Avg_RESP", "Avg_TOTAL")])

```

```{r}
# Combine row data into one dataframe for analysis
# Create a new column "Row" to indicate the distance category (Row 1, Row 2, etc.)

row_data <- rbind(
  data.frame(Distance = "Row 1", result[result$Session == 2 & result$PN %in% c(6, 12, 14, 17, 19), c("Avg_PM1", "Avg_PM2.5", "Avg_PM10", "Avg_RESP", "Avg_TOTAL")]),
  data.frame(Distance = "Row 2", result[result$Session == 2 & result$PN %in% c(5, 9, 13, 16), c("Avg_PM1", "Avg_PM2.5", "Avg_PM10", "Avg_RESP", "Avg_TOTAL")]),
  data.frame(Distance = "Row 3", result[result$Session == 2 & result$PN %in% c(2, 4, 8, 12), c("Avg_PM1", "Avg_PM2.5", "Avg_PM10", "Avg_RESP", "Avg_TOTAL")]),
  data.frame(Distance = "Row 4", result[result$Session == 2 & result$PN %in% c(1, 3, 7), c("Avg_PM1", "Avg_PM2.5", "Avg_PM10", "Avg_RESP", "Avg_TOTAL")])
)

# Run ANOVA for each parameter (PM1, PM2.5, PM10, RESP, TOTAL)
params <- c("Avg_PM1", "Avg_PM2.5", "Avg_PM10", "Avg_RESP", "Avg_TOTAL")
anova_results <- list()

for (param in params) {
  # Perform ANOVA for each parameter
  anova_result <- aov(get(param) ~ Distance, data = row_data)
  
  # Store the result
  anova_results[[param]] <- summary(anova_result)
}

# Print the results of the ANOVA tests
for (param in names(anova_results)) {
  cat(paste("ANOVA results for", param, ":\n"))
  print(anova_results[[param]])
  cat("\n")
}

```
Avg_PM1:
p-value = 0.37 (Not significant at the 0.05 level)
There is no statistically significant difference in the mean PM1 values across 
the different distances from the source.

Avg_PM2.5:
p-value = 0.359 (Not significant at the 0.05 level)
There is no statistically significant difference in the mean PM2.5 values across 
the different distances from the source.

Avg_PM10:
p-value = 0.417 (Not significant at the 0.05 level)
There is no statistically significant difference in the mean PM10 values across 
the different distances from the source.

Avg_RESP:
p-value = 0.338 (Not significant at the 0.05 level)
There is no statistically significant difference in the mean RESP values across 
the different distances from the source.

Avg_TOTAL:
p-value = 0.48 (Not significant at the 0.05 level)
There is no statistically significant difference in the mean TOTAL values across the different distances from the source.

# Is there significance difference Between buildings & vegetation 
```{r}


```
